\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-07 paper/main.tex
Spell-check

\begin{figure}[H]
    \centering
    \includegraphics[width=0.975\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments monitoring
    the movement of Dendra2 proteins photoconverted by laser targeting ROI 1. 
    Bold lines are mean values of the cohort ($n=8$) and shaded regions cover
    one standard devation errors in the mean.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    \textbf{d.} Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions. Scale bars 50 \textmu m.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.80, 0.76, 0.86, 0.60 and 0.38

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard devation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

When combining all three modifications,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while the mean concentrations are reduced in both the anterior
and the posterior ventricles (ROIs 5 and 6).
To summarize, these results show that the ventricular geometry and the
cross-sectional area of the ventricular ducts clearly impact solute
transport, with notable differences in transport resulting from modest
geometry modifications. These findings may explain the larger
differences in the cilia-deficient computational model results
compared to the differences in experimental data in \Cref{fig:fig4},
as the computational geometry is not an exact representation of the
individual zebrafish ventricles.

@@

The Brezzi-Douglas-Marini and discontinuous Galerkin scheme
$\mathrm{BDM}_1$-$\mathrm{DG}_0$
we used to discretize the Stokes equations
is computationally more expensive
than \fixme{conventional methods} \miro{ the more commonly used discretizations} such as a
Taylor-Hood ($P_2-P_{1}$ elements)~\cite{Stenberg1990ErrorProblem} or lower-order enriched
elements (e.g. MINI)~\cite{Brezzi2011MixedMethods}.
Here, $P_k$ is the space of continuous Lagrange polynomials of degree less
than or equal to the integer $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property;
by the definition of the elements, the numerical approximation of
the CSF velocity $\uu_h$ satisfies the continuity equation (\Cref{eq:stokes_eqs_cont}).
Resultingly $\normltwo{\nabla\cdot\uu_h}=0$ to within solver tolerances,
an important property for the stability of the advection-diffusion
equation, especially in advection-dominated cases.
\cite{cesmelioglu2022compatible, johnson2009numerical, gresho1998incompressible}
Additionally, the BDM-DG scheme 
lets us impose the impermeability condition $\uu\cdot\nn=0$
strongly, since the BDM elements control the normal component of the 
velocity~\cite{Brezzi1985TwoProblems}.
For elements where this is not the case (such as Taylor-Hood elements),
the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers,
Bertoluzza2017BoundaryHemodynamics}, which affects
convergence properties \miro{$<<$what do you mean here?} and increases problem complexity
(see \Cref{subsec:appendixA1} for more details).

@@

Convergence rates reported in~\Cref{subsec:numerical_verification} were
approximated in the following way. Define the error $e_k$ for a given mesh
resolution $h_k$. Based on the assumption that the error scales as
$e_k \approx C{h_k}^\eta$, where $C$ is a constant independent of $h_k$,
the convergence order $\eta$ was approximated as
\begin{equation}
    \eta = \frac{\log{(e_k/e_{k+1}})}{\log{(h_k/h_{k+1}})}.
    \label{eq:convergence_order_estimate}
\end{equation}
\miro{
We define the approximation of the infinity norm of a finite element function
$\phi = \phi_i\psi_i(\xx)$
as 
\begin{equation*}
    \normlinf{\phi} = \max_i{\vert \phi_i\vert}.
\end{equation*}
Here $\psi_i$ are the basis functions and $\phi_i$ are the degrees of freedom defined as point envaluations}

@@

The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme that we employ for the Stokes equations
is computationally more expensive than a more conventional approach based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($P_2-P_1$ elements) or lower-order enriched
elements~\cite{Brezzi2011MixedMethods},
such as the MINI-element.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
fundamental to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. Preliminarily, we attempted
to discretize the Stokes equations with 
Taylor-Hood elements, but with the resulting numerical approximation
of the velocity $\uu_h$, we were unable to achieve
stability of the advection-diffusion discretization.
Numerical experiments identified non-zero divergence of $\uu_h$
as the root cause of these stability issues.
On the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity.
On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$
remained practically unchanged. This is a result
of the fact that incompressibility is only achieved in a weak sense when
employing Taylor-Hood elements. The advective part of the total flux
is
\begin{equation*}
    \nabla\cdot(c\uu) = c(\nabla\cdot\uu) + \uu\cdot\nabla c.
\end{equation*}
In the case that $\nabla\cdot\uu\neq 0$, a source (or sink) term for the concentration
is introduced. This may introduce errors in the numerical approximation of $c$
and lead to instabilities and spurious oscillations, especially in the case of
advection-dominated transport
\cite{gresho1998incompressible, cesmelioglu2022compatible, johnson2009numerical}.
The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme avoids these issues,
since $\nabla\cdot\uu_h=0$ is satisfied exactly as a result of the element definitions.

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=0.975\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments monitoring
    the movement of Dendra2 proteins photoconverted by laser targeting ROI 1. 
    Bold lines are mean values of the cohort ($n=8$) and shaded regions cover
    one standard deviation errors in the mean.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    \textbf{d.} Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions. Scale bars 50 \textmu m.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.80, 0.76, 0.86, 0.60 and 0.38

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard deviation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

When combining all three modifications,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while the mean concentrations are reduced in both the anterior
and the posterior ventricles (ROIs 5 and 6).
To summarize, these results show that the ventricular geometry and the
cross-sectional area of the ventricular ducts clearly impact solute
transport, with notable differences in transport resulting from modest
geometry modifications. These findings may also explain the larger
differences in the cilia-deficient computational model results
compared to the differences in experimental data in \Cref{fig:fig4},
as the computational geometry is not an exact representation of the
individual zebrafish ventricles.

@@

The Brezzi-Douglas-Marini and discontinuous Galerkin scheme
$\mathrm{BDM}_1$-$\mathrm{DG}_0$
we used to discretize the Stokes equations
is computationally more expensive
than \fixme{conventional methods} \miro{ the more commonly used discretizations} such as a
Taylor-Hood ($P_2-P_{1}$ elements)~\cite{Stenberg1990ErrorProblem} or lower-order enriched
elements (e.g. MINI)~\cite{Brezzi2011MixedMethods}.
Here, $P_k$ is the space of continuous Lagrange polynomials of degree less
than or equal to the integer $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property;
by the definition of the elements, the numerical approximation of
the CSF velocity $\uu_h$ satisfies the continuity equation (\Cref{eq:stokes_eqs_cont}).
As a result $\normltwo{\nabla\cdot\uu_h}=0$ to within solver tolerances,
an important property for the stability of the advection-diffusion
equation, especially in advection-dominated cases.
\cite{cesmelioglu2022compatible, johnson2009numerical, gresho1998incompressible}
Additionally, the BDM-DG scheme 
lets us impose the impermeability condition $\uu\cdot\nn=0$
strongly, since the BDM elements control the normal component of the 
velocity~\cite{Brezzi1985TwoProblems}.
For elements where this is not the case (such as Taylor-Hood elements),
the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers,
Bertoluzza2017BoundaryHemodynamics}, which affects
convergence properties \miro{$<<$what do you mean here?} and increases problem complexity
(see \Cref{subsec:appendixA1} for more details).

@@

Convergence rates reported in~\Cref{subsec:numerical_verification} were
approximated in the following way. Define the error $e_k$ for a given mesh
resolution $h_k$. Based on the assumption that the error scales as
$e_k \approx C{h_k}^\eta$, where $C$ is a constant independent of $h_k$,
the convergence order $\eta$ was approximated as
\begin{equation}
    \eta = \frac{\log{(e_k/e_{k+1}})}{\log{(h_k/h_{k+1}})}.
    \label{eq:convergence_order_estimate}
\end{equation}
\miro{
We define the approximation of the infinity norm of a finite element function
$\phi = \phi_i\psi_i(\xx)$
as 
\begin{equation*}
    \normlinf{\phi} = \max_i{\vert \phi_i\vert}.
\end{equation*}
Here $\psi_i$ are the basis functions and $\phi_i$ are the degrees of freedom defined as point evaluations}

@@

The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme that we employ for the Stokes equations
is computationally more expensive than a more conventional approach based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($P_2-P_1$ elements) or lower-order enriched
elements~\cite{Brezzi2011MixedMethods},
such as the MINI-element.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
fundamental to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. First, we attempted
to discretize the Stokes equations with 
Taylor-Hood elements, but with the resulting numerical approximation
of the velocity $\uu_h$, we were unable to achieve
stability of the advection-diffusion discretization.
Numerical experiments identified non-zero divergence of $\uu_h$
as the root cause of these stability issues.
On the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity.
On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$
remained practically unchanged. This is a result
of the fact that incompressibility is only achieved in a weak sense when
employing Taylor-Hood elements. The advective part of the total flux
is
\begin{equation*}
    \nabla\cdot(c\uu) = c(\nabla\cdot\uu) + \uu\cdot\nabla c.
\end{equation*}
In the case that $\nabla\cdot\uu\neq 0$, a source (or sink) term for the concentration
is introduced. This may introduce errors in the numerical approximation of $c$
and lead to instabilities and spurious oscillations, especially in the case of
advection-dominated transport
\cite{gresho1998incompressible, cesmelioglu2022compatible, johnson2009numerical}.
The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme avoids these issues,
since $\nabla\cdot\uu_h=0$ is satisfied exactly as a result of the element definitions.
